The Recursive Bernstein-Vazirani Algorithm

### 2018-06-01 Quantum Computation/rBernsteinVazirani.tex
update rBV to make it clear.
about the qubit registers

The Hadamard gate can be written in the following notation:

\begin{equation}
  H = \frac{1}{\sqrt{2^n}}\sum_{\vec{x},\vec{y}} (-1)^{\vec{x}\vdot\vec{y}}
  \ket{\vec{x}}\bra{\vec{y}}.
\end{equation}

Define $\ket{P_{\vec{x}}}$ as
\begin{equation}
  \ket{P_{\vec{x}}} = \frac{1}{\sqrt{2^n}} \sum_{\vec{y}}
  (-1)^{\vec{x}\vdot\vec{y}} \ket{\vec{y}}.
\end{equation}
One can verify that
\begin{equation}
  \braket{P_{\vec{x}} |P_{\vec{y}}} = \delta_{\vec{x},\vec{y}},\,
  \sum_{\vec{x}}\ketbra{P_{\vec{x}}} = \id.
\end{equation}
And the Hadamard gate could be written as
\begin{equation}
  H = \sum_{\vec{x}} \ket{\vec{x}}\bra{P_{\vec{x}}} 
  = \sum_{\vec{x}} \ket{P_{\vec{x}}} \bra{\vec{x}}.
\end{equation}
Therefore the Hadamard gate transfer between the basis where information is
recorded in qubit register $\ket{\vec{x}}$ and the basis where information is
recorded in the phase $\ket{P_{\vec{x}}}$. A special state which will be used
often is
\begin{equation}
  \ket{P_0} = \frac{1}{\sqrt{2^n}} \sum_{\vec{x}}\ket{\vec{x}}.
\end{equation}

@@

Define the quantum version of oracles as
\begin{align}
  & f\ket{\vec{x}_1,\cdots,\vec{x}_k} =
  (-1)^{\vec{x}_k\vdot\vec{s}_{\vec{x}_1,\cdots,\vec{x}_{k-1}}} 
  \ket{\vec{x}_1,\cdots,\vec{x}_k},\\
  & g_j \ket{\vec{x}_1,\cdots, \vec{x}_j, 
    \vec{s}_{\vec{x}_1,\cdots,\vec{x}_j} ,\cdots,\vec{x}_k}\nonumber\\
  & = (-1)^{\vec{x}_j \vdot \vec{s}_{\vec{x}_1,\cdots,\vec{x}_{j-1}}}
    \ket{\vec{x}_1,\cdots, \vec{x}_j, 
    \vec{s}_{\vec{x}_1,\cdots,\vec{x}_j} ,\cdots,\vec{x}_k},\nonumber\\
  & \;\text{for } j=k-1,k-2,\cdots ,2,\\
  & g_1 \ket{\vec{x}_1, \vec{s}_{\vec{x}_1}, \vec{x}_3, \cdots,\vec{x}_k} =
  (-1)^{\vec{x}_1 \vdot \vec{s}} \ket{\vec{x}_1, \vec{s}_{\vec{x}_1}, \vec{x}_3, \cdots,\vec{x}_k},
\end{align}
where we have used the phase kickback trick mentioned in section~\ref{sec:Phase
kick} and ignored the ancilla qubits. Notice that all quantum oracles are
meaningless phase gate unless a superposition of basis
$\ket{\vec{x}_1,\cdots,\vec{x}_k}$ are present. In the following, we will always
ensure a superposition of the form 
\[
  \sum_{\vec{x}_1,\cdots,\vec{x}_j} \phi(\vec{x}_1,\cdots,\vec{x}_j,\text{(something)})
  \ket{\vec{x}_1,\cdots,\vec{x}_j,\text{(something)}},
\]
is present for $g_j$ to act on. Here $\phi$ are some non-zero phases.

A quantum algorithm to solve rBV problem depends on the following operators
$D_j(j=1,\cdots,k)$, defined recursively by:

@@

Here the $H^{\otimes n}_j$ is the $n$ bit Hadamard gate $H^{\otimes n}$ acting
on the $j$th register. Each operator $D_{k-j}$ solve the $j$th level rBV
problem. Specifically, the first operator $D_k$ does the following:
\begin{align}
  & D_k \sum_{\vec{x}_1,\cdots,\vec{x}_k} \ket{\vec{x}_1,\cdots,\vec{x}_k}\nonumber\\ 
  = & H^{\otimes n}_k \sum_{\vec{x}_1,\cdots,\vec{x}_k}
  (-1)^{\vec{x}_k\vdot\vec{s}_{\vec{x}_1,\cdots,\vec{x}_{k-1}}} 
  \ket{\vec{x}_1,\cdots,\vec{x}_k} \\
  = & H_k^{\otimes n} \sum_{\vec{x}_1,\cdots,\vec{x}_{k-1}}
  \ket{\vec{x}_1,\cdots,\vec{x}_{k-1},P_{\vec{s}_{\vec{x}_1,\cdots,\vec{x}_{k-1}}}}
  \\
  = & \sum_{\vec{x}_1,\cdots,\vec{x}_{k-1}}
  \ket{\vec{x}_1,\cdots,\vec{x}_{k-1},\vec{s}_{\vec{x}_1,\cdots,\vec{x}_{k-1}}}.
\end{align}
Notice that if there are some arbitrary non-zero phases
$\phi(\vec{x}_1,\cdots,\vec{x}_k)$ in front of
$\ket{\vec{x}_1,\cdots,\vec{x}_k}$ on the first line, the final result will not
change except from this arbitrary phase. $D_k$ can be understood as the following
diagram:
\begin{equation}
\begin{tikzcd}
  \ket{\cdots,\vec{x}_k} \arrow[r, bend left, "D_k"] & 
   \arrow[l, bend left, "D^\dagger_k"] 
   \ket{\cdots,\vec{s}_{\vec{x}_1,\cdots,\vec{x}_{k-1}}}.
\end{tikzcd}
\end{equation}

The second operator $D_{k-1}$ does the following next:
\begin{align}
  & D_{k-1} \sum_{\vec{x}_1,\cdots,\vec{x}_{k-1}}
  \ket{\vec{x}_1,\cdots,\vec{x}_{k-1},\vec{s}_{\vec{x}_1,\cdots,\vec{x}_{k-1}}}
  \\
  =& H^{\otimes n}_{k-1} D^\dagger_k \sum_{\vec{x}_1,\cdots,\vec{x}_{k-1}} 
  (-1)^{\vec{x}_{k-1}\vdot \vec{s}_{\vec{x}_1,\cdots,\vec{x}_{k-2}}}
  \ket{\vec{x}_1,\cdots,\vec{x}_{k-1},\vec{s}_{\vec{x}_1,\cdots,\vec{x}_{k-1}}}
  \\
  =& H^{\otimes n}_{k-1} \sum_{\vec{x}_1,\cdots,\vec{x}_{k}}
  (-1)^{\vec{x}_{k-1}\vdot \vec{s}_{\vec{x}_1,\cdots,\vec{x}_{k-2}}}
  \ket{\vec{x}_1,\cdots,\vec{x}_{k-1},\vec{x}_{k}}
  \\
  =& H^{\otimes n}_{k-1} \sum_{\vec{x}_1,\cdots,\vec{x}_{k-2}}
  \ket{\vec{x}_1,\cdots,\vec{x}_{k-1},
  P_{\vec{s}_{\vec{x}_1,\cdots,\vec{x}_{k-2}}},P_0}
  \\
  =& \sum_{\vec{x}_1,\cdots,\vec{x}_{k-2}}
  \ket{\vec{x}_1,\cdots,\vec{x}_{k-2}, \vec{s}_{\vec{x}_1,\cdots,\vec{x}_{k-2}},P_0}
\end{align}

@@

\subsection{Complexity of quantum algorithm}
\label{sec:Complexity of quantum algorithm}
The number of steps taken in $D_k$ is $2$. In $D_{k-1}$ is $2+2=2^2$, in
$D_{k-2}$ is $2+2^2+2=2^3$. In general, in $D_{k-j}$ is
\begin{equation}
  2^1 + 2^2 + \cdots + 2^{j-1} + 2 = 2^{j+1}
\end{equation}

Hence, the total complexity is:
\begin{equation}
  2 + 2^2 + \cdots + 2^k = 2^{k+1}-2 \approx O(2^k)
\end{equation}

>>>>

The Hadamard gate on $n$ qubits $H^{\otimes n}$ can be written in the following notation:

\begin{equation}
  H^{\otimes n} = \frac{1}{\sqrt{2^n}}\sum_{\vec{x},\vec{y}} (-1)^{\vec{x}\vdot\vec{y}}
  \ket{\vec{x}}\bra{\vec{y}}.
\end{equation}

Define $\ket{P_{\vec{x}}}$ as
\begin{equation}
  \ket{P_{\vec{x}}} = \frac{1}{\sqrt{2^n}} \sum_{\vec{y}}
  (-1)^{\vec{x}\vdot\vec{y}} \ket{\vec{y}}.
\end{equation}
One can verify that
\begin{equation}
  \braket{P_{\vec{x}} |P_{\vec{y}}} = \delta_{\vec{x},\vec{y}},\,
  \sum_{\vec{x}}\ketbra{P_{\vec{x}}} = \id.
\end{equation}
And the Hadamard gate could be written as
\begin{equation}
  H^{\otimes n} = \sum_{\vec{x}} \ket{\vec{x}}\bra{P_{\vec{x}}} 
  = \sum_{\vec{x}} \ket{P_{\vec{x}}} \bra{\vec{x}}.
\end{equation}
Therefore the Hadamard gate transfer between the basis where information is
recorded in qubit register ($\ket{\vec{x}}$) and the basis where information is
recorded in the phase ($\ket{P_{\vec{x}}}$). A special state which will be used
often is
\begin{equation}
  \ket{P_0} = \frac{1}{\sqrt{2^n}} \sum_{\vec{x}}\ket{\vec{x}}.
\end{equation}

@@

Here we present a simple quantum algorithm to the $k$th level rBV problem. The
quantum register are initialised in the state:
\begin{equation}
  \ket{P_0}^{\otimes k}\otimes \ket{-} \otimes \cdots \otimes \ket{-} 
  = \left(\frac{1}{\sqrt{2^n}}\right)^k
  \sum_{\vec{x}_1,\cdots,\vec{x}_{k}} \ket{\vec{x}_1,\cdots,\vec{x}_k}
  \ket{-} \otimes \cdots \otimes \ket{-}.
\end{equation}

Here the first $k\times n$ qubits are grouped into $k$ qubit registers, each
having $n$ qubits. The rest of qubits are ancilla qubits prepared in the
$\ket{-} = (\ket{0}-\ket{1})/\sqrt{2} $ state. Although for simplicity we will
ignore them in the following calculation, it should be noted that the ancilla
qubits are important for the phase kickback trick mentioned before. Also for
simplicity, we will ignore all normalisation constant for for the following
discussion. Therefore the initial state of qubit register is written as:
\begin{equation}
  \sum_{\vec{x}_1,\cdots,\vec{x}_{k}} \ket{\vec{x}_1,\cdots,\vec{x}_k}
\end{equation}

Define the quantum version of oracles as
\begin{align}
  & f\ket{\vec{x}_1,\cdots,\vec{x}_k} =
  (-1)^{\vec{x}_k\vdot\vec{s}_{\vec{x}_1,\cdots,\vec{x}_{k-1}}} 
  \ket{\vec{x}_1,\cdots,\vec{x}_k},\\
  & g_j \ket{\vec{x}_1,\cdots, \vec{x}_j, 
  \vec{s}_{\vec{x}_1,\cdots,\vec{x}_j} ,\vec{x}_{j+2},\cdots,\vec{x}_k}\nonumber\\
  & = (-1)^{\vec{x}_j \vdot \vec{s}_{\vec{x}_1,\cdots,\vec{x}_{j-1}}}
    \ket{\vec{x}_1,\cdots, \vec{x}_j, 
    \vec{s}_{\vec{x}_1,\cdots,\vec{x}_j} ,\vec{x}_{j+2},\cdots,\vec{x}_k},\nonumber\\
  & \;\text{for } j=k-1,k-2,\cdots ,2,\\
  & g_1 \ket{\vec{x}_1, \vec{s}_{\vec{x}_1}, \vec{x}_3, \cdots,\vec{x}_k} =
  (-1)^{\vec{x}_1 \vdot \vec{s}} \ket{\vec{x}_1, \vec{s}_{\vec{x}_1}, \vec{x}_3, \cdots,\vec{x}_k},
\end{align}
where we have used the phase kickback trick mentioned in section~\ref{sec:Phase
kick} and ignored the ancilla qubits. Notice that all quantum oracles are
meaningless phase gate unless a superposition of basis
$\ket{\vec{x}_1,\cdots,\vec{x}_k}$ are present. In the following, we will always
ensure a superposition of the form 
\[
  \sum_{\vec{x}_1,\cdots,\vec{x}_j} \phi(\vec{x}_1,\cdots,\vec{x}_j,\text{(something)})
  \ket{\vec{x}_1,\cdots,\vec{x}_j,\text{(something)}},
\]
is present for $g_j$ to act on. Here $\phi$ are some non-zero phases.

A quantum algorithm to solve the rBV problem depends on the following operators
$D_j(j=1,\cdots,k)$, defined recursively by:

@@

Here the $H^{\otimes n}_j$ is simply the $n$ bit Hadamard gate $H^{\otimes n}$ acting
on the $j$th register. Each operator $D_{k-j}$ solve the $j$th level rBV
problem. Specifically, the first operator $D_k$ does the following:
\begin{align}
  & D_k \sum_{\vec{x}_1,\cdots,\vec{x}_k} \ket{\vec{x}_1,\cdots,\vec{x}_k}\nonumber\\ 
  = & H^{\otimes n}_k \sum_{\vec{x}_1,\cdots,\vec{x}_k}
  (-1)^{\vec{x}_k\vdot\vec{s}_{\vec{x}_1,\cdots,\vec{x}_{k-1}}} 
  \ket{\vec{x}_1,\cdots,\vec{x}_k} \\
  = & H_k^{\otimes n} \sum_{\vec{x}_1,\cdots,\vec{x}_{k-1}}
  \ket{\vec{x}_1,\cdots,\vec{x}_{k-1},P_{\vec{s}_{\vec{x}_1,\cdots,\vec{x}_{k-1}}}}
  \\
  = & \sum_{\vec{x}_1,\cdots,\vec{x}_{k-1}}
  \ket{\vec{x}_1,\cdots,\vec{x}_{k-1},\vec{s}_{\vec{x}_1,\cdots,\vec{x}_{k-1}}}.
\end{align}
Notice that if there are some arbitrary non-zero phases
$\phi(\vec{x}_1,\cdots,\vec{x}_k)$ in front of
$\ket{\vec{x}_1,\cdots,\vec{x}_k}$ on the first line, the final result will not
change except from this arbitrary phase. $D_k$ can be understood as the following
diagram:
\begin{equation}
\begin{tikzcd}
  \ket{\cdots,\vec{x}_k} \arrow[r, bend left, "D_k"] & 
   \arrow[l, bend left, "D^\dagger_k"] 
   \ket{\cdots,\vec{s}_{\vec{x}_1,\cdots,\vec{x}_{k-1}}}.
\end{tikzcd}
\end{equation}

The second operator $D_{k-1}$ does the following next:
\begin{align}
  & D_{k-1} \sum_{\vec{x}_1,\cdots,\vec{x}_{k-1}}
  \ket{\vec{x}_1,\cdots,\vec{x}_{k-1},\vec{s}_{\vec{x}_1,\cdots,\vec{x}_{k-1}}}
  \\
  =& H^{\otimes n}_{k-1} D^\dagger_k \sum_{\vec{x}_1,\cdots,\vec{x}_{k-1}} 
  (-1)^{\vec{x}_{k-1}\vdot \vec{s}_{\vec{x}_1,\cdots,\vec{x}_{k-2}}}
  \ket{\vec{x}_1,\cdots,\vec{x}_{k-1},\vec{s}_{\vec{x}_1,\cdots,\vec{x}_{k-1}}}
  \\
  =& H^{\otimes n}_{k-1} \sum_{\vec{x}_1,\cdots,\vec{x}_{k}}
  (-1)^{\vec{x}_{k-1}\vdot \vec{s}_{\vec{x}_1,\cdots,\vec{x}_{k-2}}}
  \ket{\vec{x}_1,\cdots,\vec{x}_{k-1},\vec{x}_{k}}
  \\
  =& H^{\otimes n}_{k-1} \sum_{\vec{x}_1,\cdots,\vec{x}_{k-2}}
  \ket{\vec{x}_1,\cdots,\vec{x}_{k-2},
  P_{\vec{s}_{\vec{x}_1,\cdots,\vec{x}_{k-2}}},P_0}
  \\
  =& \sum_{\vec{x}_1,\cdots,\vec{x}_{k-2}}
  \ket{\vec{x}_1,\cdots,\vec{x}_{k-2}, \vec{s}_{\vec{x}_1,\cdots,\vec{x}_{k-2}},P_0}
\end{align}

@@

\subsection{Complexity of quantum algorithm}
\label{sec:Complexity of quantum algorithm}
The number of gates ($n$ bit Hadamard and quantum queries) taken in $D_k$ is
$2$. In $D_{k-1}$ is $2+2=2^2$, in $D_{k-2}$ is $2+2^2+2=2^3$. In general, in
$D_{k-j}$ there are
\begin{equation}
  2^1 + 2^2 + \cdots + 2^{j-1} + 2 = 2^{j+1},
\end{equation}
$n$ bit Hadamard gates and quantum query gates. Hence, the total complexity is:
\begin{equation}
  2 + 2^2 + \cdots + 2^k = 2^{k+1}-2 \approx O(2^k)
\end{equation}
